---
title: 'Visualizing Data Part 3: Scatterplots'
author: "Chris Kozak"
output: html_document
---

```{r message=FALSE}
library(Lahman) #a comprehensive dataset of MLB statistics
library(nflscrapR) #contains NFL data and a built-in API for scraping the most recent data
library(dplyr) #package for querying data
library(ggplot2) #one of the best packages for data visualization
library(ggiraph) #extension that makes ggplot2 graphs interactive
```


##Scatterplots

Next, we'll create some scatterplots using both base R and ggplot to visualize both baseball and football data. As usual, we'll begin by filtering the data with dplyr. The following query will select the career homeruns and career strikeouts of all players, and filter it down to only those players with more than 400 homeruns.

```{r}
bdat1d <- Batting%>%
  group_by(playerID)%>%
  summarize(career_HR=sum(HR),career_SO=sum(SO))%>%
  filter(career_HR>=400)

head(bdat1d)
```
  
At this point, we have the playerIDs, but they list, for example, Hank Aaron as "aaronha01" and Jeff Bagwell as "bagweje01." We want a dataframe featuring these players with their actual names. To get that, we'll need to use the following **inner_join** command from dplyr to merge our new data frame with the names from the Master table.

```{r}
HR_vs_SO <- inner_join(bdat1d,Master,by=c("playerID"))%>% #this has the bdat1d dataframe and Lahman Master dataframe do an inner_join on playerID
  select(nameFirst,nameLast,career_HR,career_SO) #this selects only the first name, last name, career homeruns, and career strikeouts from this table, to create the new HR_vs_SO dataframe

head(HR_vs_SO)
```

Finally, we'll use the paste command to put the first name and last name together into a new column.

```{r}
HR_vs_SO$name <- paste(HR_vs_SO$nameFirst,HR_vs_SO$nameLast)

head(HR_vs_SO)
```

Next, we'll move on to the fun part: visualizations. Here's a scatterplot generated using base R:

```{r}
plot(HR_vs_SO$career_SO, HR_vs_SO$career_HR, xlab="career strikeouts", ylab="career homeruns", main="Career Homeruns vs Strikeouts for Great Hitters")
```

And the same scatterplot generated using ggplot2:

```{r}
g<-ggplot()+
  geom_point_interactive(data=HR_vs_SO,aes(x=career_SO, y=career_HR, tooltip=name,data_id=nameLast))+
  ggtitle("Career Homeruns vs Strikeouts for Great Hitters")+
  xlab("career strikeouts")+
  ylab("career homeruns")

ggiraph(code = print(g), hover_css = "fill:white;stroke:black")
```

With the ggplot scatterplot, we can hover over each datapoint and see, for example, that Barry Bonds has the most career homeruns (762), and a respectably small number of strikeouts (1539). Now, we'll do two similar scatterplots featuring NFL passing touchdowns vs interceptions for the 2016 season.

```{r}
NFLq1 <- playerstats16%>%
  group_by(name)%>%
  summarize(season_TDs=sum(pass.tds),season_INTs=sum(pass.ints))%>%
  filter(season_TDs>=1)

plot(NFLq1$season_TDs, NFLq1$season_INTs, main="Passing Touchdowns vs Interceptions (2016)", xlab="2016 Passing TDs", ylab="2016 INTs")

g<-ggplot()+
  geom_point_interactive(data=NFLq1,aes(x=season_TDs, y=season_INTs, tooltip=name))+
  ggtitle("Passing Touchdowns vs Interceptions (2016)")+
  xlab("2016 Passing TDs")+
  ylab("2016 INTs")

ggiraph(code = print(g), hover_css = "fill:white;stroke:black")

```

In our next post, we'll review bar charts.